
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
  keep_md: true
---

------

## Loading and preprocessing the data

```{r, echo = FALSE, results = "hide"}
options(scipen = 999) # Disable scientific notation
Sys.setlocale("LC_ALL","English") # Set english for outputs
setwd("C:/Users/Pepe/Desktop/Peer Assessment 1") # My working directory
```

1. Loading the data
```{r}
repdata <- read.csv("activity.csv")
```

2. Transforming "date" variable to R `'Date'` class
```{r}
repdata$date <- as.Date(repdata$date, format = "%Y-%m-%d")
```

------

## What is mean total number of steps taken per day?

1. Histogram of the total number of steps taken each day

+ **First**: calculating the **mean of total number of steps per day** (using `aggregate` function)
```{r}
steps_per_day <- aggregate(steps ~ date, data = repdata, FUN = sum)
```

+ **Second**: histogram
```{r}
hist(
       steps_per_day$steps,
       main = "Histogram", 
       xlab = "Number of steps per interval"
)
```

2. Reporting the **mean** and **median** of total number of steps per day
```{r}
Mean <- mean(steps_per_day$steps)
Median <- median(steps_per_day$steps)
```
+ The **mean** and **median** of total number of steps taken per day are `r round(Mean, 3)` and `r round(Median, 3)`, respectively.

------

## What is the average daily activity pattern?

1. Time series plot.
+ **First**: Calculating the average daily activity pattern
```{r}
daily_activity_pattern <- aggregate(
       steps ~ interval, 
       data = repdata, 
       FUN = mean
)
```

+ **Second**: The plot
```{r}
plot(
       steps ~ interval, 
       data = daily_activity_pattern, 
       ylab = "Number of steps",
       type = "l"
)
```

2. The 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is:
```{r}
daily_activity_pattern[which.max(daily_activity_pattern$steps), ]
```


------

## Imputing missing values

1. Total number of missing values in the dataset (i.e. the total number of rows with NAs):
```{r}
sum(!complete.cases(repdata))
```

2. Strategy for filling in all of the missing values in the dataset: the mean value for a 5-minute interval was used.

3. New dataset (`repdata_filled`) with the missing data filled in.
```{r}
repdata_filled <- data.frame() # empty data frame
intervals <- unique(repdata$interval) # 5-minute intervals
for(i in 1:length(intervals)){
       dfi <- subset(repdata, interval == intervals[i]) # i interval data
       mean_step <- mean(dfi$steps, na.rm = TRUE) # mean value for i interval 
       # imputing
       dfi$steps <- ifelse(is.na(dfi$steps), yes = mean_step, no = dfi$steps)
       repdata_filled <- rbind(repdata_filled, dfi) # filling empty data frame
}
```

4. Histogram of the total number of steps per day and mean and median of total number of steps taken per day.

+ **First**: calculating the **mean of total number of steps per day** 
```{r}
steps_per_day <- aggregate(steps ~ date, data = repdata_filled, FUN = sum)
```

+ **Second**: histogram
```{r}
hist(
       steps_per_day$steps,
       main = "Histogram", 
       xlab = "Number of steps per interval"
)
```

+ **Third**: reporting the **mean** and **median** of total number of steps per day
```{r}
Mean <- mean(steps_per_day$steps)
Median <- median(steps_per_day$steps)
```

The **mean** and **median** of total number of steps taken per day are `r round(Mean, 3)` and `r round(Median, 3)`, respectively. Even though the imputation strategy did not have an impact on the mean value, the median value was affected, which implies an effect on the variation and distribution of data.

------

## Are there differences in activity patterns between weekdays and weekends?

1. Creating a new factor variable with two levels -- "weekday" and "weekend" using the filled-in missing dataset.

+ **First**: a function called `days` is created. This function returns "weekend" if date is 'saturday' or 'sunday'. Otherwise, it returns "weekday"
```{r}
days <- function(date){
       # date: an object of class "Date".
       ifelse(
               weekdays(date) == "Sunday" | weekdays(date) == "Saturday",
               yes = "weekend", 
               no = "weekday"
       )  
}
```

+ **Second**: using the `days` function to create the new **factor** variable
```{r}
repdata_filled$weekday <- as.factor(days(repdata_filled$date))
```

2. Panel plot containing a time series plot of the 5-minute interval and the average number of steps, averaged across all weekday days or weekend days.

+ **First**: calculating the average number of steps per interval and day class (weekday or weekend):
```{r}
daily_activity_pattern <- aggregate(
       steps ~ interval*weekday, 
       data = repdata_filled, 
       FUN = mean
)
```

+ **Second**: time series plot.
```{r}
library(lattice) 
xyplot(
       steps ~ interval | weekday, 
       data = daily_activity_pattern, 
       type = "l",
       ylab = "Number of steps",
       layout = c(1, 2)
)
```